---
title: "Supplementary Data: Radiocarbon Age Modeling for Kiwulan Chronology"
author: "Li-Ying Wang and Ben Marwick"
date: "8/18/2020"
output: html_document
bibliography: references.bib
---

# Introduction

This document provides data and code for calibrating radiocarbon dates and modeling the chronology of the upper component of Kiwulan. The full discussion of chronology is in the paper: Standardization of ceramic shape: A case study from the Iron Age pottery from northeastern Taiwan. There are 17 radiocarbon dates from the upper component according to the excavation report [@Chen2007]. We calibrated 11 radiocarbon dates after removing 6 dates that are not informative due to their long time ranges without standard deviation (< 200 yr). We then modeled two phases related to the European influence, pre- and post- European periods. Our purpose is to answer whether the radiocarbon dates support a time range that corresponds to our artifact-based chronology in terms of the presence of Europeans in Taiwan. 

```{r load-data}
kw_dates <- readxl::read_excel(here::here("analysis", "data", "raw_data", "KWL_C14_dates.xlsx"))
```

```{r tidy-data}
library(tidyverse)

# subset dates with error value
kw_dates_with_error <-
  kw_dates  %>%
  filter(!is.na(se))

# convert rce to numeric
kw_dates_with_error$rce <- as.numeric(kw_dates_with_error$rce)

# make some kind of ID
kw_dates_with_error$id <- with(kw_dates_with_error,
                               paste0(Square,"-",Layer))

# upper layer with estimation of phases
kw_dates_with_error_three_phases <-
  kw_dates_with_error %>%
  filter(Component == "upper") %>%
  mutate(phase = rce) %>%
  mutate(phase = ifelse(rce > 324, "Pre European", phase),
         phase = ifelse(rce < 324, "Post European", phase)) %>%
  select(`Lab code`, "rce", "se", "phase", Context)
```

```{r edit-calibrate-function, include = FALSE}
print.oxcAARCalibratedDatesList <- function(x, ...){
  n <- length(x)
  if (n == 1) print(x[[1]],...)
  else{
    cat(sprintf("List of %d calibrated dates:\n",n))
    for (i in 1:n) {
      print(x[[i]])
    }
  }
}

#' @export
plot.oxcAARCalibratedDatesList <- function(x, use_ggplot = T, ...){
  if (length(x) == 1) plot(x[[1]], use_ggplot, ...)
  else{
    if (requireNamespace("ggplot2", quietly = TRUE) & use_ggplot) {
      plotoxcAARCalibratedDatesListGGPlot2(x, ...)
    } else {
      plotoxcAARCalibratedDatesListSystemGraphics(x, ...)
    }
  }
}

plotoxcAARCalibratedDatesListGGPlot2<-function(x, ...){
  .data <- NULL
  to_plot<-lapply(x,function(y){
    alpha <- ifelse((!(is.null(y$posterior_probabilities))& !(is.na(y$posterior_probabilities))),
                    c(0.25,0.75),
                      c(0.75,0)
                      )
    outdata <- data.frame(
      dates=y$raw_probabilities$dates,
      probability=y$raw_probabilities$probabilities,
      name=y$name,
      class = "unmodelled",
      alpha=alpha[1])
    if(!(is.null(y$posterior_probabilities)) & !(is.na(y$posterior_probabilities))) {
      outdata <- rbind(outdata,
                   data.frame(
                     dates=y$posterior_probabilities$dates,
                     probability=y$posterior_probabilities$probabilities,
                     name=y$name,
                     class = "modelled",
                     alpha=alpha[2])
      )
    }
    return(outdata)

  })
  to_plot <- do.call(rbind,to_plot)

  m <- ggplot2::ggplot(to_plot)

  graph <- m + ggridges::geom_ridgeline(ggplot2::aes(x = .data$dates,
                                  y = .data$name,
                                  height = .data$probability,
                                  alpha = .data$alpha,
                                  group = interaction(.data$class,.data$name)),
                              scale=100,
                              fill = "#fc8d62",
                              color="#00000077") +
    ggplot2::theme_light() + ggplot2::labs(y="Dates", x = "AD")  +
    ggplot2::scale_alpha_continuous(guide = FALSE)

  methods::show(graph)
}

plotoxcAARCalibratedDatesListSystemGraphics <- function(x, ...){
  op <- graphics::par(no.readonly = TRUE)
  years <- years_post <- NA

  indices <- 1:length(x)

  min_year <- min(
    sapply(
      indices,
      function(i) {
        this_year_range <- get_years_range(x[[i]])
        if (!all(is.na(this_year_range))) {
          min(this_year_range, na.rm=T)
        } else {
          NA
        }
      }
    ), na.rm=T
  )

  max_year <- max(
    sapply(
      indices,
      function(i) {
        this_year_range <- get_years_range(x[[i]])
        if (!all(is.na(this_year_range))) {
          max(this_year_range, na.rm=T)
        } else {
          NA
        }
      }
    ), na.rm=T
  )

  graphics::par(mfrow=c(length(x)+1,1))

  graphics::par(oma = c(3,1,2,2) + 0.1,
                mar = c(0,1,0,1) + 0.1)

  for (i in indices) {
    max_prob <- 0
    years <- probability <- NA
    post_present <- prob_present <- FALSE

    if(class(x[[i]]$raw_probabilities)=="data.frame") {
      prob_present <- TRUE
      years <- x[[i]]$raw_probabilities$dates
      probability <- x[[i]]$raw_probabilities$probabilities
      max_prob <- max(probability)
    }

    unmodelled_color <- "lightgrey"

    years_post <- probability_post <- NA

    if(class(x[[i]]$posterior_probabilities)=="data.frame") {
      post_present <- TRUE
      years_post <- x[[i]]$posterior_probabilities$dates
      probability_post <- x[[i]]$posterior_probabilities$probabilities
      unmodelled_color <- "#eeeeeeee"
      max_prob <- max(max_prob, probability_post)
    }

    if(!prob_present & !post_present)
    {
      year_range <-c(0,1)
    } else {
      year_range <- get_years_range(x[[i]])
    }

    graphics::plot(
      years,
      probability,
      type = "n",
      ylim=c(max_prob / 7 * -1,max_prob),
      xlim = c(min_year,max_year),
      axes = FALSE
    )
    graphics::axis(side=4)
    graphics::polygon(years, probability, border = "black", col = unmodelled_color)
    if (unmodelled_color!="lightgrey"){
      graphics::polygon(years_post, probability_post, border = "black", col = "#aaaaaaaa")
    }
    graphics::text(x=min_year, y=max_prob, labels=print_label(x[[i]]),las=2,cex=0.6, adj=0)
    graphics::grid()
  }
  plot(c(min_year,max_year),c(0,0),
       axes = FALSE,
       type="n")
  graphics::axis(side=1, pretty(c(min_year, max_year)))
  graphics::par(op)
}
```

# Calibration

We calibrated 11 dates from different contexts, including artifact-bearing deposit, burials, and middens, at Kiwulan using oxcAAR package, which executes OxCal v4.4.1 with radiocarbon calibration curve IntCal20 [@Reimer2020]. 

```{r calibrate-C14data}
# Use oxcAAR to calibrate, run the following two lines to download ocxAAR 
# library(devtools)
# install_github('ISAAKiel/oxcAAR')
library(oxcAAR)
quickSetupOxcal()
date_kwl <- oxcalCalibrate(kw_dates_with_error$rce,
                           kw_dates_with_error$se,
                           kw_dates_with_error$`Lab code`)
date_kwl
plot(date_kwl)
calcurve_plot(date_kwl)
```

# Chronological model 

We constructed a Bayesian radiocarbon calibration model for the 11 dates to obtain the range of the pre-European and the post-European periods we examined. We assume that those two phases are in a time sequence with the arrival of the Europeans in 1624 as a break [@Andrade2007]. The end of the post-European period is identified by the start of the Chinese migration to Yilan in 1780s [@Chen2007]. 

The results suggest that the pre-European period started between 998–1386 cal AD at a 95.4% probability range and ended between 1496–1656 cal AD at a 95.4% probability range. The start date of the post-European period is the same as the end date of the pre-European period (1496–1656 cal AD). The end date of the post-European period is 1524-1770 cal AD at a 93.4% probability range. In general, the overall model for radiocarbon age relative to the European contact is consistent with our artifact-based chronology.

```{r modeling-upper-layer}
# contiguous phases, not overlapping
three_phases_con <- 'Plot()
{
    Sequence("Upper Layer")
    {
      Boundary("Start of Pre-European");
      Phase()
      {
        R_Date("NTU-3791", 340, 30);
        R_Date("NTU-4292", 510, 75);
        R_Date("NTU-4310", 360, 100);
        R_Date("NTU-4320", 340, 100);
        R_Date("NTU-4304", 600, 75);
        R_Date("NTU-4423", 610, 90);
        R_Date("NTU-4319", 920, 105);
        Interval( "Duration of Pre-European");
      };
        Boundary("Pre-European - Post-European");
      Phase()
      {
        R_Date("NTU-3993", 250, 40);
        R_Date("NTU-4016", 270, 45);
        R_Date("NTU-4311", 310, 100);
        R_Date("NTU-4419", 280, 70);
        Interval("Duration of Post-European");
      };
      Boundary("End of Post-European");
    };
};
'
```

```{r plot}
three_phases_con_results <- executeOxcalScript(three_phases_con)
three_phases_con_text <- readOxcalOutput(three_phases_con_results)
three_phases_result_con_data <- parseOxcalOutput(three_phases_con_text, only.R_Date = T) 
# str(three_phases_result_con_data)
# print(three_phases_result_con_data)
plot(three_phases_result_con_data)
```
