# Hypotheses and expectations

My hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then I expect to see the changes in social organization from a more corporate to a more network organization that can be reflected by pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then I expect the local pottery will show more homogeneous features after contact due to the pottery specialization caused by control of small group of individuals.

# Part of Introduction

In this paper, we use EFA for pottery shape analysis to evaluate the level of standardization of pottery from an Iron Age site in northeastern Taiwan in relation to the European presence in the 17th century that might hint the emergence of craft specialization for pottery production. Craft specialization can provide further evidence about the type of production organization which is closely related to economic or political aspects of society [@Arnold2000; @Blackman1993; @Costin1991; @Hirshman2010; @Stark1995]. The emergence of craft specialization can also be an indicator of social inequality that hints the possibility of the presence of production groups worked for specific individuals. The presence of craft specialization is commonly studied by measuring the standardization of ceramics based on the assumption that the specialized mass production will lead to uniformity or homogeneity of the product due to increased skills, routinization, and lower diversity of producers [@Arnold2000]. To test the prediction of a shift in the level of social inequality in this case study in northeastern Taiwan influenced by the colonial European presence, we use standardization of pottery in shapes as an index for craft specialization to identify the presence of a few individuals controlling large scale pottery production [@Costin1995; @Hirshman2010] using outline-based geometric morphometrics couples with traditional metric measurements to compare and evaluate the use of outline approaches for pottery with high consistency in shapes. 

#--------------original metric data analysis------------
Table \@ref(tab:metricCVstest) shows the results of Modified signed-likelihood ratio test (MSLRT) for equality of CVs for for six measurements and two ratio values. There is a significant difference between CV values for two metric measurements across phases, the rim diameter and the neck diameter. The CV values show that the variation of rim diameter is small (6.8%)  before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%). For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%). <!-- these numeric values in this paragraph should come from inline code, not copy-pasted in like this -->
<!-- You need to put the code that computes the MSLRT results right in here, in the Rmd --> 

```{r metricCVstest}
metric_cvs_with_phase <- readr::read_csv(here::here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))
metric_cvs_with_phase_tidy <-
  metric_cvs_with_phase %>% 
  mutate_at(c("pre-e", "post-e", "ch-con", "MSLRT", "p_value"), round, 4) %>% 
  rename(`pre-European` = "pre-e", `after European` = "post-e", Chinese = "ch-con")
knitr::kable(metric_cvs_with_phase_tidy,
             caption = "Coefficient of Variation test for metric attributes by phases")
```

#------------revised metric data analysis---------------

Table \@ref(tab:metricCVstest-pre-post) and \@ref(tab:metricCVstest-post-chi) shows the results of Asymptotic test [@Feltz1996] for equality of CVs for for six measurements and two ratio values. There is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presnece. The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%). For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period (10.6%). After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference from post-European period (p-value less than 0.05).

```{r metricCVstest-pre-post, eval=FALSE}
# read metric data
kwl_p_ave <- readr::read_csv(here::here("analysis", "data", "raw_data",
                                 "KWL-pottery-metrics-tidy-ave.csv"))
library(cvequality)
# select metric attributes and tidy data for computing CV
kwl_p_metrics_long <-
  kwl_p_ave %>%
  mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
         thi_rim_body = Thick_Rim_ave/Thick_Body_ave) %>%
  select(period,
         Dia_Rim_ave,
         Dia_Neck_ave,
         Dia_Body_ave,
         Thick_Rim_ave,
         Thick_Neck_ave,
         Thick_Body_ave,
         dia_rim_body,
         thi_rim_body,
         # H_Rim
  ) %>%
  rename(`Rim diameter (mm)` = Dia_Rim_ave,
         `Neck diameter (mm)` = Dia_Neck_ave,
         `Body diameter (mm)` = Dia_Body_ave,
         `Rim thickness (mm)` = Thick_Rim_ave,
         `Neck thickness (mm)` = Thick_Neck_ave,
         `Body thickness (mm)` = Thick_Body_ave,
         `Ratio of Rim/Body diameter` = dia_rim_body,
         `Ratio of Rim/Body thickness` = thi_rim_body
  ) %>%
  gather(variable, value, -period) %>%
  filter(!is.na(period),
         !is.na(value))
# filter pre-e and post-e periods
kwl_p_metrics_long_pre_post <-
  kwl_p_metrics_long %>%
  filter(!period == "ch-con")
# filter post-e and chi periods
kwl_p_metrics_long_post_chi <-
  kwl_p_metrics_long %>%
  filter(!period == "pre-e")
#------------------------------------------CV test for pre-e and post-e pair
kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post <-
  kwl_p_metrics_long_pre_post %>%
  nest(-variable) %>%  # or nest(period, value) %>%
  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
                               ~bind_cols(asymptotic_test (
                                 .x$value,
                                 .x$period)))) %>%
  unnest(asymptotic_test)
# compute CV
kwl_p_cvs_pre_post <-
  kwl_p_metrics_long_pre_post %>%
  group_by(period, variable) %>%
  summarise(cvs = raster::cv(value, na.rm = TRUE))
# join p-values to CVs
kwl_p_cvs_and_pvalues_pre_post <-
  kwl_p_cvs_pre_post %>%
  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post) %>%
  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
  mutate(variable = factor(variable, ordered = TRUE)) %>%
  arrange(variable)
# making table for CV
kwl_p_cvs_table_pre_post <-
  kwl_p_cvs_and_pvalues_pre_post %>%
  select(-data, -significant) %>%
  pivot_wider(names_from = period, values_from = cvs) %>%
  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
                                                "Neck thickness (mm)",
                                                "Body thickness (mm)",
                                                "Ratio of Rim/Body thickness",
                                                "Rim diameter (mm)",
                                                "Neck diameter (mm)",
                                                "Body diameter (mm)",
                                                "Ratio of Rim/Body diameter"))) %>%
  arrange(variable) %>%
  select(variable, `pre-e`, `post-e`, `D_AD`, `p_value`) %>%
  mutate_at(c("pre-e", "post-e", "D_AD", "p_value"), round, 4)
knitr::kable(kwl_p_cvs_table_pre_post,
             caption = "Coefficient of Variation test for metric attributes before and after European presence")
```
knitr::kable(metric_cvs_with_phase_tidy,
             caption = "Coefficient of Variation test for metric attributes by phases")
```{r metricCVstest-post-chi, eval=FALSE}
#-----------------------------------CV test for post-e and ch-con pair
kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi <-
  kwl_p_metrics_long_post_chi %>%
  nest(-variable) %>%  # or nest(period, value) %>%
  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
                               ~bind_cols(asymptotic_test (
                                 .x$value,
                                 .x$period)))) %>%
  unnest(asymptotic_test)
# compute CV
kwl_p_cvs_pre_post <-
  kwl_p_metrics_long_post_chi %>%
  group_by(period, variable) %>%
  summarise(cvs = raster::cv(value, na.rm = TRUE))
# join p-values to CVs
kwl_p_cvs_and_pvalues_post_chi <-
  kwl_p_cvs_pre_post %>%
  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi) %>%
  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
  mutate(variable = factor(variable, ordered = TRUE)) %>%
  arrange(variable)
# making table for CV
kwl_p_cvs_table_post_chi <-
  kwl_p_cvs_and_pvalues_post_chi %>%
  select(-data, -significant) %>%
  pivot_wider(names_from = period, values_from = cvs) %>%
  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
                                                "Neck thickness (mm)",
                                                "Body thickness (mm)",
                                                "Ratio of Rim/Body thickness",
                                                "Rim diameter (mm)",
                                                "Neck diameter (mm)",
                                                "Body diameter (mm)",
                                                "Ratio of Rim/Body diameter"))) %>%
  arrange(variable) %>%
  select(variable, `post-e`, `ch-con`, `D_AD`, `p_value`) %>%
  mutate_at(c("post-e", "ch-con", "D_AD", "p_value"), round, 4)
knitr::kable(kwl_p_cvs_table_post_chi,
             caption = "Coefficient of Variation test for metric attributes between European and Chinese presence")
